---
title: 'Describing Data Part 2'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "my-theme.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#17139C",link_color = "#DD3E3E", base_font_size = "18px")

```

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(MASS) # for simulating variables with specific correlations
library(tidyverse) # for graphing
library(ggExtra) # for marginal distributions
library(psych) # for dataset showing content overlap
```
### Last time

.pull-left[### Population Variability

**Sums of squares**
$$\small SS = \Sigma(X_i-\mu_x)^2$$
**Variance**
$$\small \sigma^2 = \frac{\Sigma(X_i-\mu_x)^2}{N} = \frac{SS}{N}$$
**Standard devation**
$$\scriptsize \sigma = \sqrt{\frac{\Sigma(X_i-\mu_x)^2}{N}}= \sqrt{\frac{SS}{N}} = \sqrt{\sigma^2}$$]


.pull-right[
###Sample variability
**Sums of squares**
$$\small SS = \Sigma(X_i-\bar{X})^2$$
**Variance**
$$\small s^2 = \frac{\Sigma(X_i-\bar{X})^2}{N-1} = \frac{SS}{N-1}$$
**Standard devation**
$$\scriptsize s = \sqrt{\frac{\Sigma(X_i-\bar{X})^2}{N-1}}= \sqrt{\frac{SS}{N-1}} = \sqrt{s^2}$$]
---

## Bi-variate descriptives

### Covariation 

"Sum of the cross-products"


### Population
$$SP_{XY} =\Sigma(X_i−\mu_X)(Y_i−\mu_Y)$$

### Sample
$$ SP_{XY} =\Sigma(X_i−\bar{X})(Y_i−\bar{Y})$$


???

**What does a large, positive SP indicate?**
A positive relationship, same sign

**What does a large, negatve SP indicate?**
A negative relationship, different sign

**What does SP close to 0 indicate?**
No relationship



---

## Covariance
Sort of like the variance of two variables

### Population
$$\sigma_{XY} =\frac{\Sigma(X_i−\mu_X)(Y_i−\mu_Y)}{N}$$

### Sample
$$s_{XY} = cov_{XY} =\frac{\Sigma(X_i−\bar{X})(Y_i−\bar{Y})}{N-1}$$


---

## Covariance table


.large[
$$\mathbf{K_{XX}} = \left[\begin{array}
{rrr}
\sigma^2_X & cov_{XY} & cov_{XZ} \\
cov_{YX} & \sigma^2_Y & cov_{YZ} \\
cov_{ZX} & cov_{ZY} & \sigma^2_Z
\end{array}\right]$$
]


$cov_{xy} = cov_{yx}$ 

---
## Covariance table


.large[
$$\mathbf{K_{XX}} = \left[\begin{array}
{rrr}
\sigma^2_X & 126.5 & 5.2 \\
126.5 & \sigma^2_Y & cov_{YZ} \\
5.2 & cov_{ZY} & \sigma^2_Z
\end{array}\right]$$
]


Which variable, $Y$ or $Z$, does $X$ have greater relationship with?

--

Can't know because you don't know what units they're measured in!

---

## Correlation

- Measure of association

- How much two variables are *linearly* related

- -1 to 1

- Sign indicates direction of relationship

- Invariant to changes in mean or scaling

---

## Correlation

Pearson product moment correlation

### Population
$$\rho_{XY} = \frac{\Sigma z_Xz_Y}{N} = \frac{SP}{\sqrt{SS_X}\sqrt{SS_Y}} = \frac{\sigma_{XY}}{\sigma_X \sigma_Y}$$
### Sample
$$r_{XY} = \frac{\Sigma z_Xz_Y}{n-1} = \frac{SP}{\sqrt{SS_X}\sqrt{SS_Y}} = \frac{s_{XY}}{s_X s_Y}$$
???

**Why is it called the Pearson Product Moment correlation?**
Pearson = Karl Pearson
Product = multiply
Moment = variance is the second moment of a distribution
---

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(50, 5)
Sigma = matrix(c(.8, .5, .5, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
```

```{r, out.width='50%'}
data |>  ggplot(aes(x = x, y = y)) + geom_point(size = 3) + theme_bw()
```

What is the correlation between these two variables?

--

Correlation = `r round(cor(data$x, data$y),2)`

---
```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(10, 100)
Sigma = matrix(c(.8, -.3, -.3, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
```

```{r, out.width='50%'}
data |>  ggplot(aes(x = x, y = y)) + geom_point(size = 3) + theme_bw()
```

What is the correlation between these two variables?

--

Correlation = `r round(cor(data$x, data$y),2)`

---
```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(3, 4)
Sigma = matrix(c(.8, 0, 0, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
```

```{r, out.width='50%'}
data |>  ggplot(aes(x = x, y = y)) + geom_point(size = 3) + theme_bw()
```

What is the correlation between these two variables?

--

Correlation = `r round(cor(data$x, data$y),2)`

---

## Effect size

- Recall that *z*-scores allow us to compare across units of measure; the products of standardized scores are themselves standardized. 

- The correlation coefficient is a **standardized effect size** which can be used communicate the strength of a relationship.

- Correlations can be compared across studies, measures, constructs, time. 

- Example: the correlation between age and height among children is $r = .70$. The correlation between self- and other-ratings of extraversion is $r = .25$. 
    
---

## What is a large correlation?

--
- [Cohen (1988)](http://www.utstat.toronto.edu/~brunner/oldclass/378f16/readings/CohenPower.pdf): .1 (small), .3 (medium), .5 (large)
    - Often forgot: Cohen said only to use them when you had nothing else to go on, and has since regretted even suggesting benchmarks to begin with. 

--

- $r^2$: Proportion of variance "explained"
    - as [Ozer & Funder (2019)](https://uopsych.github.io/psy611/readings/Ozer_Funder_2019.pdf) discuss, we're not really explaining anything and the change in scale can mess up our interpretations if we're not careful.

---

## What are good benchmarks? 

From Ozer & Funder (2019)

- Classic social psych studies: $r = .36-.42$
- Scarcity increases the perceived alue of a commodity $r = .12$
- People attribute failures to bad luck $r = .10$
- Communicators perceived as more credible are more persuasive $r = .10$
- People in a bad mood are more aggressive $r = .41$
- Antihistamine and symptom relief $r = .11$
- Ibuprofen and pain relief $r = .14$
- Height and weight $r = .44$

---

## What are good benchmarks? 

**Implications**
- Don't dismiss small effects
- Be skeptical of large effects

**Recommendations**
- Report effect sizes 
- Use large samples -- remember bias?
- Report effect sizes in context
- Stop using empty terminology
- Revise guidelines

???
**What are good benchmarks?**
- Classic social psych studies: $r$ between .35 and .45
- scarcity increases the perceived value of a commodity ($r = .12$), 
- people attribute failures to bad luck ($r = .10$), 
- communicators perceived as more credible are more persuasive ($r = .10$)
- people in a bad mood are more aggressive ($r = .41$)
- SES and IQ predicting mortality $r = 25$

- Antihistamine and symptom relief: $r = .11$
- Ibuprofen and pain relief: $r = .14$
- Men weigh more than women: $r = .26$
- High elevations have lower annual temps: $r = .34$
- Height and weight $r = .44$

---

## What affects correlations?

It's not enough to calculate a correlation between two variables. You MUST always look at a figure of the data to make sure the number accurately describes the relationship. Correlations can be easily fooled by qualities of your data, like:

- Skewed distributions

- Outliers

- Restriction of range

- Nonlinearity

---

## Skewed distributions

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(3, 4)
Sigma = matrix(c(.8, .2, .2, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
data$x = data$x^4
```

```{r, fig.height = 5}
p = data |>  ggplot(aes(x=x, y=y)) + geom_point() + theme_bw(base_size = 25)
ggMarginal(p, type = "density")
```
---

## Outliers

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(3, 4)
Sigma = matrix(c(.8, 0, 0, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 50, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
data[51, ] = c(7, 10)
```

```{r, fig.height = 5}
data |>  ggplot(aes(x=x, y=y)) + geom_point() + theme_bw(base_size = 25)
```
---

## Outliers

.pull-left-small[
.code-small[
```{r, eval=FALSE, message = FALSE, warning = FALSE}
data |>  ggplot(aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "red") +
  geom_smooth(data = data[-51,], 
              method = "lm", 
              se = FALSE) +
  theme_bw(base_size = 25)
```
]
]

.pull-right-large[
```{r, echo=FALSE, message = FALSE, warning = FALSE}
data |>  ggplot(aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "red") +
  geom_smooth(data = data[-51,], 
              method = "lm", 
              se = FALSE) +
  theme_bw(base_size = 25)
```
]

---

## Restriction of range

```{r, echo = FALSE}
set.seed(1010191) # so we all get the same random numbers
mu = c(100, 4)
Sigma = matrix(c(.7, .4, 4, .75), ncol = 2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
real_data = data
data = filter(data, x >100 & x < 101)
```

.pull-left-small[
.code-small[
```{r, eval = FALSE, message = FALSE, warning = FALSE}
data |> 
ggplot(aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "red") +
  theme_bw(base_size = 25)
```
]
]

.pull-right-large[
```{r, echo = FALSE, message = FALSE, warning = FALSE}
data |>  ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") + theme_bw(base_size = 25)
```
]

???

What if I told you there were scores on X could range from 97 to 103?

---

## Restriction of range

.pull-left-small[
.code-small[

```{r, eval = FALSE, warning=F, message=F}
data |> 
ggplot(aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "red") + 
  geom_point(data = real_data) +
  geom_smooth(method = "lm",
              se = FALSE,
              data = real_data,
              color = "blue") +
  theme_bw(base_size = 25)
```
]
]

.pull-right-large[
```{r, echo = FALSE, warning=F, message=F}
data |> 
ggplot(aes(x=x, y=y)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "red") + 
  geom_point(data = real_data) +
  geom_smooth(method = "lm",
              se = FALSE,
              data = real_data,
              color = "blue") +
  theme_bw(base_size = 25)
```
]



???

**Can you think of example where this might occur in psychology?**
My idea: that many psychology studies only look at undergraduates (restricted age, restricted education) -- can't use these as predictors or covariates

---

## Nonlinearity

```{r, echo = FALSE, warning = FALSE, message=FALSE}
x = runif(n = 150, min = -2, max = 2)
y = x^2 +rnorm(n = 150, sd = .5)
data = data.frame(x,y)
```

```{r, fig.height = 5, warning=FALSE, message=FALSE, echo=FALSE}
data |>  ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") + theme_bw(base_size = 25)
```

---

## It's not always apparent

Sometimes issues that affect correlations won't appear in your graph, but you still need to know how to look for them.

- Low reliability

- Content overlap

- Multiple groups

---

## Multiple groups

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
m_mu = c(100, 4)
m_Sigma = matrix(c(.7, .4, 4, .75), ncol = 2) #diagonals are reliabilites, off-diagonals are correlations
m_data = mvrnorm(n = 150, mu = m_mu, Sigma = m_Sigma)
m_data = as.data.frame(m_data)
colnames(m_data) = c("x", "y")

f_mu = c(102, 3)
f_Sigma = matrix(c(.7, .4, 4, .75), ncol = 2) #diagonals are reliabilites, off-diagonals are correlations
f_data = mvrnorm(n = 150, mu = f_mu, Sigma = f_Sigma)
f_data = as.data.frame(f_data)
colnames(f_data) = c("x", "y")

m_data$gender = "male"
f_data$gender = "female"

data = rbind(m_data, f_data)
```

```{r, fig.height = 5, message=F, warning=F, echo=FALSE}
data %>% ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") + theme_bw(base_size = 25)
```

---

## Multiple groups

```{r, fig.height = 5, message=F, warning=F, echo=FALSE}
data %>% ggplot(aes(x=x, y=y, color = gender)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + guides(color = F) + theme_bw(base_size = 25)
```

--

Known as **Simpson's Paradox**
---

### Special cases of the Pearson correlation

- **Spearman/Kendall correlation coefficient**
    - Applies when both X and Y are ranks (ordinal data) instead of continuous
    - Spearman for larger samples, Kendall for smaller samples (or a lot of ties in rank ordering). 

- **Point-biserial correlation coefficient**
    - Applies when Y is binary.
        - NOTE: This is not an appropriate statistic when you [artificially dichotomize data](../readings/Cohen_1983.pdf).

- **Phi (
$\phi$
) coefficient**
  - Both X and Y are dichotomous.


---
## Do the special cases matter?

For Spearman, you'll get a different answer. 
```{r}
x = rnorm(n = 10); y = rnorm(n = 10) #randomly generate 10 numbers from normal distribution
```

.pull-left[
```{r, echo = T}
head(cbind(x,y))
cor(x,y, method = "pearson")
```
]
.pull-right[
```{r}
head(cbind(x,y, rank(x), rank(y)))
cor(x,y, method = "spearman")
```
]

---
## Do the special cases matter?

If your data are naturally binary, no difference between Pearson and point-biserial.

```{r}
x = rnorm(n = 10); y = rbinom(n = 10, size = 1, prob = .3)
head(cbind(x,y))
```

.pull-left[
```{r, echo = T}
cor(x,y, method = "pearson")
```
]
.pull-right[
```{r}
ltm::biserial.cor(x,y, level = 2)
```
]
---
## Do the special cases matter?

If your data are artificially binary, there can be big differences.

```{r}
x = rnorm(n = 10); y = rnorm(n = 10)
```

.pull-left[
```{r, echo = T}
head(cbind(x,y))
cor(x,y, method = "pearson")
```
]
.pull-right[
```{r}
d_y = ifelse(y < median(y), 0, 1)
head(cbind(x,y, d_y))
ltm::biserial.cor(x,d_y, level = 2)
```
]

### Don't use median splits!
---
class: inverse

## Next time...

Probability